---
title: "208_Tutorial"
author: "Megan, Justin, Katie E"
date: "1/29/2019"
output: html_document
---


```{r setup, include=}
datasets::iris
?iris
head(iris)
summary(iris)
iris
```


1. General data work with iris (QHP):
  -data exploration: shaprio test, qqplot 
  -t-test
  -regression
  -bar graph vs. boxplot vs. piechart (boo)
2. shannon-diversity calculation
  -maybe a figure? (provide a simple data set- clean csv/excel file, a template of sorts)
3. survivorship/mortality work
  -figure




We can also use R to investigate and explore other datasets, as well as answer different types of questions.


We can also use R to investigate and explore other datasets, as well as answer different types of questions. For example, we can use R to investigate diversity. One measure of diversity we learn about in 208 is the Shannon Diversity index. Conveniently, there is a package in R (vegan) specifically designed to calculate not only the Shannon Diversity index, but to also investigate other measures of diversity.



```{r Shannon-diversity index}

#first need to install vegan

#load vegan 
library(vegan)

#Using a preloaded dataset BCI
data("BCI")

#look at BCI dataset
BCI

#investigate this dataset
ncol(BCI)
colnames(BCI)
nrow(BCI)
rownames(BCI)

#Two different ways to get shannon-diversity index for each site
diversity(BCI)

#This second method is more complex, but allows us to choose other possible indices to use instead of Shannon. Here we are going to assign our calculation to a new dataset labelled diversity_dataset
diversity_dataset <- diversity(BCI, index= "shannon", MARGIN = 1, base = exp(1))
diversity_dataset

#checking the normality of this data
qqnorm(diversity_dataset)
qqline(diversity_dataset)

#using a histogram to look at data
hist(diversity_dataset)

#investigate diversity_dataset
plot(diversity_dataset)
summary(diversity_dataset)

#getting the median diversity
median(diversity_dataset)

#getting the average diversity
mean(diversity_dataset)

#Creating a species accumulation curve. Species accumulation curves allow us to visualize how many species are present for a certain number of sites. Among other functions, it allows us to determine if species count is increasing with sampling effort (number of sites).
species_accumulation <- specaccum(BCI)

#Plotting a basic species accumulation curve
plot(species_accumulation) 

#We can edit this graph to better visualize the data, and to be generally more visually appealing 
#Aesthetics include:
#ci.type = bar, line, or polygon; Type of confidence interval drawn on the figure. 
#col=""; color of species accumulation line
#lwd= ; width of line
#ci.lty= 0,1,2,3... ; Type of border on confidence polygon or confidence lines
#ci.col=""; color for confidence line or polygon
#ylab= "" ; this allows us to edit the label on the y axis
plot(species_accumulation, ci.type="poly", col="blue", lwd=3, ci.lty=0, ci.col="orange", ylab ="Species Accumulation") 
```

```{r J}
if (!require("ggplot2")) install.packages("ggplot2"); library(ggplot2)
if (!require("UsingR")) install.packages("UsingR"); library(UsingR)
```

```{r setup, include=TRUE}
library(ggplot2)
datasets::iris
head(iris)
tail(iris)
summary(iris)

#print out the whole dataset
iris

ggplot(iris, aes(x=Sepal.Length, y =Petal.Length))+
  geom_point(aes(colour = factor(Species)), shape =1, size=3)+
  theme_classic() +
  geom_smooth(method=lm, color="blue", se=TRUE)+
  xlab("Sepal Length")+
  ylab("Petal Length")+
  ylim(0,8)+
  labs(color = "Species")

  #xlim(0,8) not sure if we want to include this change as it makes the graph a little goofy looking

#http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/
```





  
---------- KE









-------


- upload data from an excel file 
  - install readxl
  
```{r Upload Data }
install.packages("readxl")
library(readxl)
demographyData <- read_excel("BIO208_Demog_Calcs_Master.xlsx")
View(demographyData)

```


- subset into groups based on year and sex

```{r Subsets}
fpre1840 <- subset(demographyData, select = "Pre-1840 Female")
mpre1840 <- subset(demographyData, select = "Pre-1840 Male")

```


- calculate proportion alive for each age class


```{r Functions}

#function that calculates the number of people still alive
numStillAlive <- function(subset1) { 
  totalStartAlive <- sum(subset1)
  for (i in row(subset1)){
    sumDead <- sum(subset1[1:i,1])
    stillAlive <- totalStartAlive - sumDead
    listStillAlive <- c(listStillAlive,stillAlive)
  }
  return(listStillAlive)
}

indvStillAlive <- as.data.frame(numStillAlive(fpre1840))
print(indvStillAlive)

proportionAlive <- function(data.frame) {
  for (i in col(indvStillAlive)) {
    a <- indvStillAlive[i]/indvStillAlive[1]
    list <- c(list,a)
  }
  return(list)
}

variable <- proportionAlive(indvStillAlive)



indvStillAlive[1]
indvStillAlive[4]/indvStillAlive[1]

#function that finds the proportion of people still alive
propStillAlive <- function(list) {
  for (i in row(list)) {
    propAlive <- list[i]/list[1]
    listOfPropAlive <- c(listOfPropAlive,propAlive)
  }
  return(listPropAlive)
}

peepStillAlive <- propStillAlive(indvStillAlive)
print(propAStillAlive)


```


```{r Calculate Proportions}
fpre1840alive <- numStillAlive(fpre1840)
fpre1840prop <- proportionAlive(fpre1840alive)

mpre1840alive <- as.data.frame(numStillAlive(mpre1840))
mpre1840prop <- proportionAlive(mpre1840alive)


```



- plot survivorship curve

```{r Plot Survivorship}

