---
title: "208_Tutorial"
author: "Megan, Justin, Katie E"
date: "1/29/2019"
output: html_document
---

<<<<<<< HEAD
```{r setup, include=}
datasets::iris
?iris
head(iris)
summary(iris)
iris
```
=======


>>>>>>> 3a4acfd784e20c5cf0b0e2b33a583e678264cde4

1. General data work with iris (QHP):
  -data exploration: shaprio test, qqplot 
  -t-test
  -regression
  -bar graph vs. boxplot vs. piechart (boo)
2. shannon-diversity calculation
  -maybe a figure? (provide a simple data set- clean csv/excel file, a template of sorts)
3. survivorship/mortality work
  -figure

<<<<<<< HEAD
We can also use R to investigate and explore other datasets, as well as answer different types of questions. For example, we can use R to investigate diversity. One measure of diversity we learn about in 208 is the Shannon Diversity index. Conveniently, there is a package in R (vegan) specifically designed to calculate not only the Shannon Diversity index, but to also investigate other measures of diversity.

```{r Shannon-diversity index}

#first need to install vegan

#load vegan 
library(vegan)

#Using a preloaded dataset BCI
data("BCI")

#look at BCI dataset
BCI

#investigate this dataset
ncol(BCI)
colnames(BCI)
nrow(BCI)
rownames(BCI)


#Two different ways to get shannon-diversity index for each site
diversity(BCI)
diversity_dataset <- diversity(BCI, index= "shannon", MARGIN = 1, base = exp(1))
diversity_dataset

#investigate diversity_dataset
plot(diversity_dataset)
summary(diversity_dataset)

#getting the median diversity
median(diversity_dataset)

#getting the average diversity
mean(diversity_dataset)

#Creating a species accumulation curve
species_accumulation <- specaccum(BCI)
plot(species_accumulation) 
plot(species_accumulation, ci.type="poly", col="red", lwd=2, ci.lty=0, ci.col="lightblue", ylab ="Species Accumulation" ) 
=======
```{r J}
if (!require("ggplot2")) install.packages("ggplot2"); library(ggplot2)
if (!require("UsingR")) install.packages("UsingR"); library(UsingR)
```

```{r setup, include=TRUE}
datasets::iris
head(iris)
tail(iris)
summary(iris)
```



```{rJ}
>>>>>>> 3a4acfd784e20c5cf0b0e2b33a583e678264cde4

```

