---
title: "208_Tutorial"
author: "Megan, Justin, Katie E"
date: "1/29/2019"
output: html_document
---


```{r setup, include=}
datasets::iris
?iris
head(iris)
summary(iris)
iris
```


1. General data work with iris (QHP):
  -data exploration: shaprio test, qqplot 
  -t-test
  -regression
  -bar graph vs. boxplot vs. piechart (boo)
2. shannon-diversity calculation
  -maybe a figure? (provide a simple data set- clean csv/excel file, a template of sorts)
3. survivorship/mortality work
  -figure




We can also use R to investigate and explore other datasets, as well as answer different types of questions.


We can also use R to investigate and explore other datasets, as well as answer different types of questions. For example, we can use R to investigate diversity. One measure of diversity we learn about in 208 is the Shannon Diversity index. Conveniently, there is a package in R (vegan) specifically designed to calculate not only the Shannon Diversity index, but to also investigate other measures of diversity.



```{r Shannon-diversity index}

#first need to install vegan

#load vegan 
library(vegan)

#Using a preloaded dataset BCI
data("BCI")

#look at BCI dataset
BCI

#investigate this dataset
ncol(BCI)
colnames(BCI)
nrow(BCI)
rownames(BCI)

#Two different ways to get shannon-diversity index for each site
diversity(BCI)

#This second method is more complex, but allows us to choose other possible indices to use instead of Shannon. Here we are going to assign our calculation to a new dataset labelled diversity_dataset
diversity_dataset <- diversity(BCI, index= "shannon", MARGIN = 1, base = exp(1))
diversity_dataset

#checking the normality of this data
qqnorm(diversity_dataset)
qqline(diversity_dataset)

#using a histogram to look at data
hist(diversity_dataset)

#investigate diversity_dataset
plot(diversity_dataset)
summary(diversity_dataset)

#getting the median diversity
median(diversity_dataset)

#getting the average diversity
mean(diversity_dataset)

#Creating a species accumulation curve. Species accumulation curves allow us to visualize how many species are present for a certain number of sites. Among other functions, it allows us to determine if species count is increasing with sampling effort (number of sites).
species_accumulation <- specaccum(BCI)

#Plotting a basic species accumulation curve
plot(species_accumulation) 

#We can edit this graph to better visualize the data, and to be generally more visually appealing 
#Aesthetics include:
#ci.type = bar, line, or polygon; Type of confidence interval drawn on the figure. 
#col=""; color of species accumulation line
#lwd= ; width of line
#ci.lty= 0,1,2,3... ; Type of border on confidence polygon or confidence lines
#ci.col=""; color for confidence line or polygon
#ylab= "" ; this allows us to edit the label on the y axis
plot(species_accumulation, ci.type="poly", col="blue", lwd=3, ci.lty=0, ci.col="orange", ylab ="Species Accumulation") 
```

```{r J}
if (!require("ggplot2")) install.packages("ggplot2"); library(ggplot2)
if (!require("UsingR")) install.packages("UsingR"); library(UsingR)
```

```{r setup, include=TRUE}
library(ggplot2)
datasets::iris
head(iris)
tail(iris)
summary(iris)

#print out the whole dataset
iris

ggplot(iris, aes(x=Sepal.Length, y =Petal.Length))+
  geom_point(aes(colour = factor(Species)), shape =1, size=3)+
  theme_classic() +
  geom_smooth(method=lm, color="blue", se=TRUE)+
  xlab("Sepal Length")+
  ylab("Petal Length")+
  ylim(0,8)+
  labs(color = "Species")

  #xlim(0,8) not sure if we want to include this change as it makes the graph a little goofy looking

#http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/
```





  
---------- KE

```{r}
## Load survival package
library(survival)
## List datasets in survival package
data(package = "survival")

## Load lung data
data(lung)

## Show first 6 rows
head(lung)
```

```{r}
## Add survival object
lung$SurvObj <- with(lung, Surv(time, status == 2))

## Check data
head(lung)
```


```{r}
## Kaplan-Meier estimator. The "log-log" confidence interval is preferred.
km.as.one <- survfit(SurvObj ~ 1, data = lung, conf.type = "log-log")
km.by.sex <- survfit(SurvObj ~ sex, data = lung, conf.type = "log-log")

## Show object
km.as.one
survfit(formula = SurvObj ~ 1, data = lung, conf.type = "log-log")

km.by.sex
survfit(formula = SurvObj ~ sex, data = lung, conf.type = "log-log")

## See survival estimates at given time (lots of outputs)
## summary(km.as.one)
## summary(km.by.sex)

## Plotting without any specification
plot(km.as.one)
plot(km.by.sex)
```

```{r}
## Load rms package
install.packages(rms)
library(rms)

## without specification
survplot(km.by.sex)
```

```{r}
## Without dots
survplot(fit  = km.by.sex,
         conf = c("none","bands","bars")[1],
         xlab = "", ylab = "Survival",
         ## xlim(0,100),
         label.curves = TRUE,                     # label curves directly
         ## label.curves = list(keys = "lines"),  # legend instead of direct label
         levels.only  = FALSE,                    # show only levels, no label
         abbrev.label = FALSE,                    # if label used, abbreviate
         ## fun = function(x) {1 - x},            # Cumulative probability plot         
         loglog   = FALSE,                        # log(-log Survival) plot
         logt     = FALSE,                        # log time
         time.inc = 100,                          # time increment
         dots     = FALSE,                         # dot grid
         n.risk   = TRUE,                         # show number at risk
         ## srt.n.risk = 0, sep.n.risk = 0.056, adj.n.risk = 1,
         ## y.n.risk = 0, cex.n.risk = 0.6
         )


#https://rstudio-pubs-static.s3.amazonaws.com/5588_72eb65bfbe0a4cb7b655d2eee0751584.html
```




-------


- upload data from an excel file 
  - install readxl
  
```{r Upload Data }
install.packages("readxl")
library(readxl)
library(ggplot2)
library(cowplot)


demographyData <- read_excel("BIO208_Demog_Calcs_Master.xlsx")
View(demographyData)

```


- subset into groups based on year and sex

```{r Subsets}
fpre1840 <- subset(demographyData, select = "Pre-1840 Female")
mpre1840 <- subset(demographyData, select = "Pre-1840 Male")

```


- calculate proportion alive for each age class


```{r Functions}

#function that calculates the number of people still alive
numStillAlive <- function(subset1) { 
  totalStartAlive <- sum(subset1)
  for (i in row(subset1)){
    sumDead <- sum(subset1[1:i,1])
    stillAlive <- totalStartAlive - sumDead
    listStillAlive <- c(listStillAlive,stillAlive)
  }
  return(listStillAlive)
}


proportionAlive <- function(dataframe) {
  for (i in col(dataframe)) {
    a <- dataframe[i]/dataframe[1]
    list <- c(list,a)
  }
  return(list)
}


```


```{r Calculate Proportions}
fpre1840alive <- as.data.frame(numStillAlive(fpre1840))
fpre1840prop <- proportionAlive(fpre1840alive)
fpre1840$PropAlive <- fpre1840prop[2:21]
fpre1840$xAxis <- 1:20

mpre1840alive <- as.data.frame(numStillAlive(mpre1840))
mpre1840prop <- proportionAlive(mpre1840alive)
mpre1840$PropAlive <- mpre1840prop[2:21]
mpre1840$xAxis <- ("0-5" "6-10" "11-15" "16-20" "21-25" "26-30" "31-35" "36-40" "41-45" "46-50" "51-55" "56-60" "61-65" "66-70" "71-75""76-80" "81-85" "86-90" "91-95" "96+")
mpre1840$xAxis <- (0-5 6-10 11-15)


```


- plot survivorship curve

```{r Plot Survivorship}
ggplot(fpre1840) +
  aes(x = as.numeric(xAxis), y = as.numeric(PropAlive)) +
  geom_line() +
  theme_cowplot() +
  xlab("Days with enough sleep to feel rested (last 7 days)") + 
  ylab("Survivorship")

```