---
title: "208_Tutorial"
author: "Megan, Justin, Katie E"
date: "1/29/2019"
output: html_document
---

```{r setup, include=FALSE}
datasets::iris
?iris
head(iris)
summary(iris)
iris
```

1. General data work with iris (QHP):
  -data exploration: shaprio test, qqplot 
  -t-test
  -regression
  -bar graph vs. boxplot vs. piechart (boo)
2. shannon-diversity calculation
  -maybe a figure? (provide a simple data set- clean csv/excel file, a template of sorts)
3. survivorship/mortality work
  -figure
  
  
  
  
---------- KE

- upload data from an excel file 
  - install readxl
  
```{r Upload Data }
install.packages("readxl")
library(readxl)
demographyData <- read_excel("BIO208_Demog_Calcs_Master.xlsx")
View(demographyData)

```




- subset into groups based on year and sex

```{r Subsets}
fp1840 <- subset(demographyData,"Pre-1840 Female" != "NA")

```


- calculate proportion alive for each age class

```{r Calculate Proportions}

```


- plot survivorship curve

```{r Plot Survivorship}

```

- test differences in life expectancy between men and women

```{r Test Life Expectancy}

```

